---
title: "Nonparametric_estimation_survival_function"
format: html
editor: visual
---

## ================================================================

## Advanced Health Data Analysis: Survival Analysis, Course 2025/26

## 23.2.2026: Nonparametric estimation of the survival function

## ================================================================

```{r}
load("Ahda_RLab1.RData")
ls()

```

## The data set ---------------------------------------------------------------

## ----------------------------------------------------------------------------

```{r}
str(remission)
Hmisc::Label(remission)
head(remission)
```

```{r}
summary(remission)
```

## Quick data exploration -----------------------------------------------------

## ----------------------------------------------------------------------------

## Number of relapses per group

```{r}
with(remission, table(treat, cens))
```

makes us assume possibility 1

## Mean follow-up time per group

```{r}
with(remission, tapply(time, treat, summary))
```

Mean follow-up time for 6MP is larger and looks like the treatment was beneficial but we need a closer look at censored events.

## Kaplan-Meier estimation of S(t) with R -------------------------------------

## ----------------------------------------------------------------------------

```{r}
library(survival)
```

## The Surv object

## ---------------

```{r}
with(remission, Surv(time, cens))
```

+ indicates right censored observation

```{r}
srem <- with(remission, Surv(time, cens))
srem
str(srem)
```

## The Kaplan-Meier estimate is obtained with the function survfit()

## -----------------------------------------------------------------

```{r}
survfit(srem ~ 1) #~1 to use whole sample
```

-   42 samples, 30 died, 42-30=12 are rigth censored
-   median survival time with 95% interval
-   see estimation of median

## More information is shown with the function summary.survfit()

## -------------------------------------------------------------

```{r}
svf <- survfit(srem ~ 1)
summary(svf)
```

-   time = uncensored event times
-   n risk = people at risk
-   point estimation of survival probability at that time point
-   first censored observation between week 6 and 7
-   after week 23: still 5 censored observations left
-   function stays constant upt to week 35 (times of censored observations)

## Some options of the function summary.survfit():

```{r}
summary(svf, censored=TRUE)
```

-   from week 23 on: no events but consored observations
-   function is not defined beyond week 35


when I am only interested in the survival probability monthly:

```{r}
summary(svf, times = seq(4, 24, 4)) #rannge, steps
```

```{r}
summary(svf, scale = 4)
```


```{r}
survfit(srem ~ 1)
```

## Estimation of the restricted mean survival time (often not necessary for basic info)

```{r}
print(svf, rmean = "common") #up to week 35
print(svf, rmean = 50) #extend to week 50 cuase it is still positive before, ASSUMING maximal survival time
```

## Information saved in the survfit and summary.survfit objects

## ------------------------------------------------------------

```{r}
names(svf)
str(svf)
names(summary(svf))
str(summary(svf))
```

## Treatment comparison with respect to S(t)

## -----------------------------------------

```{r}
(svf2t <- survfit(srem ~ treat, remission))
summary(svf2t)
```

-   compared control and treatment group
-   control: estimated survival function goes down to zero, all died
- No upper limit for median of tratment group cause the intervalls in the bottom table do not drop below 0,5

## Graphical representation of the estimated survival functions ---------------

## ----------------------------------------------------------------------------

```{r}
plot(svf)
```

## A little bit nicer

```{r}
#windows(width = 10)
par(font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(5, 5, 4, 2))
plot(svf, lwd = 3, xlab = "Weeks", ylab = expression(bolditalic(hat(S)(t))))
title("Survival function of the time to leukemia relapse")
```

## Both treatment arms

## -------------------

windows(width = 10)

```{r}
par(font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(5, 5, 4, 2))
plot(svf2t, col = 3:4, conf.int = TRUE, lwd = 3, xlab = "Weeks", bty = "n",
     ylab = expression(bolditalic(hat(S)(t))), xaxs = "i", yaxs = "i")
legend("topright", levels(remission$treat), col = 3:4, lwd = 3, bty = "n")
```

## Maybe better without the confidence intervals

```{r}
plot(svf2t, col = 3:4, xlab = "Time to relapse [Weeks]",
     ylab = expression(bolditalic(hat(S)(t))),
     lty = 1:2, lwd = 3, yaxs = "i", xaxs = "i", bty = "n")
title("Survival functions according to treatment")
legend("bottomleft", levels(remission$treat), title = "Treatment",
       bty = "n", col = 3:4, lty = 1:2, lwd = 3)
```

## Representation of the distribution function (cumulative incidence)

## ------------------------------------------------------------------

windows(width = 10)

```{r}
par(font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(5, 5, 4, 2))
plot(svf2t, col = 3:4, xlab = "Time to relapse [Weeks]",
     ylab = expression(bolditalic(hat(F)(t))), xaxs = "i",
     lty = 1:2, fun = "event", lwd = 3, yaxs = "i", bty = "n") #define fun ="event"
title("Distribution functions according to treatment")
legend("topleft", levels(remission$treat), title = "Treatment",
       bty = "n", col = 3:4, lty = 1:2, lwd = 3)
```

## Representation of the cumulative hazard function

## ------------------------------------------------

windows(width = 10) par(font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(5, 5, 4, 2)) plot(svf2t, col = 3:4, xlab = "Time to relapse \[Weeks\]", ylab = expression(bolditalic(hat(Lambda)(t))), xaxs = "i", lty = 1:2, fun = ..., lwd = 3, yaxs = "i", bty = "l") title("Cumulative hazards functions according to treatment") legend("topleft", levels(remission\$treat), title = "Treatment", bty = "n", col = 3:4, lty = 1:2, lwd = 3)
